Modelling the emergence of whisker barrels

### 2019-12-13 paper/briefpaper_stuedits/brief.tex
A few minor edits to paper

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^l.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($l>1$) with the density of axon branching. Axon branching is modelled as 
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{./MainFig.png}
\end{center}

@@

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 52 pairs of $\gamma$ values were sampled, at the coordinates of 52 barreloid centers recreated from  \cite{haidarliu_size_2001}.

A cortical boundary enclosing 52 corresponding barrels was traced from a cytochrome oxidase stain from  \cite{shimogori_fibroblast_2005}, and Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=52$ projections on the resulting domain, using $M=2$ linear signalling gradients aligned with the anterior-posterior and  medial-lateral axes. From random initial conditions for $a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged. A reduction in the Honda delta metric (see \cite{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged within $\approx 3000$ iterations (see Figure 1). The sum of squared differences between the real and simulated barrel centres revealed a strong correspondence.

To further investigate the interplay of genetic and intrinsic mechanisms we simulated a seminal barrel duplication experiment \cite{shimogori_fibroblast_2005},  first by solving on a cortical domain comprising two separate, identically shaped cortical boundaries. In this case one distinct cortical field per thalamic projection emerged, within one of the two boundaries only, i.e., no duplication. However, merging the two fields to create an extended mirror-symmetric boundary shape, with an anterior-posterior guidance field reflected at the join, gave rise to two mirror-symmetrical barrel fields comprising 104 barrels, i.e., two cortical fields for each thalamic projection (see Figure 2). Together these results suggest that by the misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005} effectively created one large barrel field rather than two distinct fields.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eqs.~\ref{eq:comp}), and iii) the branch density of each projection is conserved (Eq.~\ref{eq:norm}).

>>>>

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a reaction-diffusion style model of how extrinsic signalling gradients can constrain the emergence of distinct fields from intrinsic cortical dynamics. Their model defines how the density of connections $c(x,t)$ and axon branches $a(x,t)$ interact at time $t$, along a 1D anterior-posterior axis $x$, for $N$ thalamocortical projections indexed by $i$. The model was derived from the assumption that the rates at which $a_i$ and $c_i$ grow are reciprocally coupled. Extending the original 1D model to simulate arealization on a 2D cortical sheet, we use $a_i(\mb{x},t)$ and $c_i(\mb{x},t)$, and  model synaptogenesis as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{k=1}^{N} c_{k}\right)[a_i]^k.
\end{equation}
%
Accordingly, where the total density of synaptic connections sums to one, connections decay at rate $\alpha$. Otherwise the connection density increases non-linearly ($k>1$) with the density of axon branching. Axon branching is modelled as
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi_i.
\end{equation}
%
The first term on the right describes the divergence (indicated by $\nabla\vcdot$) of the quantity in parentheses, which is referred to as the `flux' of axonal branching. The flux represents diffusion across the cortical sheet, at rate $D$, and the influence of $M$ molecular signalling fields, $\rho(\mb{x})$. The influence of a given field (indexed by $j$) on a given thalamic projection (indexed by $i$), is determined by $\gamma_{i,j}$, which may be positive or negative in order that axons may branch in the direction of either higher or lower  concentrations. The second term on the right represents an assumption that axon branching decreases when synaptogenesis increases, and vice versa, i.e., that axon branches are required for synaptic connections to form. Here $\chi_i=0$ is a placeholder.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{../briefpaper/MainFig.png}
\end{center}

@@

The only differences between thalamic projections are their strengths of interaction with the guidance fields, $\gamma$, hence any reliable differences between emergent cortical fields must be due to differences in these values only. We speculate that a projection interacts with a given ephrin field in the cortical subplate with a strength determined by the concentration of a similar molecule at its thalamic origin, i.e., the putative barreloid center. As such, two orthogonal linear thalamic gradients were defined, from which 41 pairs of $\gamma$ values were sampled, at the coordinates of 41 barreloid centers recreated from \cite{haidarliu_size_2001}.

A cortical boundary enclosing 41 corresponding barrels was traced from a cytochrome oxidase stain from \cite{zheng_signal_2001}, and Eqs.~\ref{eq:dc}--\ref{eq:norm} were solved for $N=41$ projections on the resulting domain, using $M=2$ linear signalling gradients aligned with the anterior-posterior and  medial-lateral axes. From random initial conditions for $a({\bf x},0)$ and $c({\bf x},0)$, a clear Voronoi-like tesselation emerged. A reduction in the Honda delta metric (see \cite{senft_mouse_1991}) confirmed that a `good' Voronoi pattern emerged within $\approx 3000$ iterations (see Figure 1). The sum of squared differences between the real and simulated barrel centres revealed a strong correspondence.

To further investigate the interplay of genetic and intrinsic mechanisms we simulated a seminal barrel duplication experiment \cite{shimogori_fibroblast_2005},  first by solving on a cortical domain comprising two separate, identically shaped cortical boundaries. In this case one distinct cortical field per thalamic projection emerged, within one of the two boundaries only, i.e., no duplication. However, merging the two fields to create an extended mirror-symmetric boundary shape, with an anterior-posterior guidance field reflected at the join, gave rise to two mirror-symmetrical barrel fields comprising NN barrels, i.e., two cortical fields for each thalamic projection (see Figure 2). Together these results suggest that by the misexpression of Fgf8, Shimogori \& Grove \citep{shimogori_fibroblast_2005} effectively created one large barrel field rather than two distinct fields.

@@

The present results suggest that the key requirements for the emergence of realistic barrel patterning are i) at each cortical location thalamocortical projections compete for a limited number of available synaptic connections (Eqs.~\ref{eq:dc}--\ref{eq:da}), ii) at each location the branching rate of a given projection is reduced by the density of other projections (Eq.~\ref{eq:comp}), and iii) the branch density of each projection is conserved (Eq.~\ref{eq:norm}).
